---
title: "Machine Learning"
author: "Ravi Kumar Tiwari"
date: "14 June 2016"
output: pdf_document
---


```{r, echo = TRUE, results='hide', message=FALSE, warning=FALSE}
library(caret)
library(rpart.plot)
library(rattle)
library(calibrate)
```

## Decision Tree Example

```{r}
head(iris)
```

Create data partition
```{r}
inTrain <- createDataPartition(iris$Species, p = 0.6, list = FALSE)
trainData <- iris[inTrain,]
testData <- iris[-inTrain,]
```

Build a decision tree model and use it for prediction on test data set
```{r}
treeModel <- train(Species ~ ., data = trainData, method = "rpart")
preClass <- predict(treeModel, newdata = testData)
cMatrix <- confusionMatrix(preClass, testData$Species)
cMatrix$table
```

Look at what are the important variables
```{r}
varImp(treeModel)
```

Visualization of the decision tree

```{r}
fancyRpartPlot(treeModel$finalModel)  
```

Alternative visualization of the decision tree

```{r, echo=FALSE}
preClassType <- ifelse(trainData$Species == "setosa", 15, 
                   ifelse(trainData$Species== "versicolor", 17, 19))
plot(trainData$Petal.Length, trainData$Petal.Width, col = trainData$Species, 
     pch = preClassType, xlab = "Petal Length", ylab = "Petal Width")
abline(v = 2.4)
segments(2.4, 1.7, 7.2, 1.7 )
text(x = 2.2, y = 1.1, labels = "Decision Boundary 1", srt = 90)
text(x = 3.9, y = 1.8, labels = "Decision Boundary 2")
```


Visualization of the prediction result

```{r, echo = FALSE}
predResult <- ifelse(preClass == testData$Species, "green", "red")

preClass <- ifelse(testData$Species == "setosa", 15, 
                   ifelse(testData$Species== "versicolor", 17, 19))


plot(testData$Petal.Length, testData$Petal.Width, col = predResult, 
     pch = preClass, cex = 1.2, xlab = "Petal Length", ylab = "Petal Width")
abline(v = 2.4)
segments(2.4, 1.7, 7.2, 1.7 )
text(x = 2.2, y = 1.1, labels = "Decision Boundary 1", srt = 90)
text(x = 3.9, y = 1.8, labels = "Decision Boundary 2")
```



## Clustering Example
### K-means clustering

```{r, echo = FALSE}
x <- seq(from=1,to=10,length.out = 10)
set.seed(10)
y1 <- rnorm(5, 4, 0.5)
set.seed(20)
y2 <- rnorm(5, 8, 0.5)
y <- c(y1,y2)
obs <- data.frame(x,y)
```

```{r}
head(obs)
```


```{r, echo = FALSE}
par(mfrow=c(1,1))
plot(x,y, cex = 1.5, pch = 19)
```

Using data to create 2 clusters
```{r}
kmeansObj <-  kmeans(obs, centers = 2)
data.frame(obs, cluster = kmeansObj$cluster)
kmeansObj$centers
```

```{r, echo=FALSE}
plot(x,y, col= kmeansObj$cluster, pch = 19, cex = 1.5, xlim = c(0,12), ylim = c(0,12))
points(kmeansObj$centers, pch = 8, cex = 2.0, col = rownames(kmeansObj$centers))
textxy(x+0.1,y + 0.1, labs = 1:10, cex = 1, col = kmeansObj$cluster)
```


Using data to create 4 clusters
```{r}
kmeansObj <-  kmeans(obs, centers = 4)
data.frame(obs, cluster = kmeansObj$cluster)
kmeansObj$centers
```


```{r, echo = FALSE}
par(mfrow=c(1,1))
plot(x,y, col= kmeansObj$cluster, pch = 19, cex = 1.5, xlim = c(0,12), ylim = c(0,12))
points(kmeansObj$centers, pch = 8, cex = 2.0, col = rownames(kmeansObj$centers))
textxy(x+0.1,y + 0.1, labs = 1:10, cex = 1, col = kmeansObj$cluster)
```


### Hierarchical Clustering
```{r, echo = FALSE}
x <- seq(from=1,to=10,length.out = 10)
set.seed(10)
y1 <- rnorm(5, 4, 0.5)
set.seed(20)
y2 <- rnorm(5, 8, 0.5)
y <- c(y1,y2)
par(mfrow=c(1,1))
plot(x,y, xlim = c(0,10), ylim = c(0,10), pch = 19)
```

```{r}
distM <- dist(obs)
clusters <- hclust(distM, method = "ward.D")
```

Create 2 partitions

```{r}
mem <- cutree(clusters, k =2)
data.frame(obs, cluster = mem)
```

```{r, echo=FALSE} 
par(mfrow=c(1,2))

plot(clusters)
abline(h=10)
text(5, 12, labels = "2 partitions")
plot(x,y, col=mem, pch = 19, cex = 1.5, xlim = c(0,12), ylim=c(0,12),
     main="Same color points form a group")
textxy(x+0.1,y + 0.1, labs = 1:10, cex = 1, col = mem)
```

Create 4 partitions

```{r}
mem <- cutree(clusters, k =4)
data.frame(obs, cluster = mem)
```

```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(clusters)
abline(h=3.5)
text(5, 5, labels = "4 partitions")
plot(x,y, col=mem, pch = 19, cex = 1.5, xlim = c(0,12), ylim=c(0,12),
     main="Same color points form a group")
textxy(x+0.1,y + 0.1, labs = 1:10, cex = 1, col = mem)
```

## Cross-validation
```{r, echo = FALSE}
drawPoly <- function(x,y, col){
  polygon(c(x,x+1,x+1,x, x), c(y,y,y+0.5,y+0.5,y), col = col)
}


plot(1, type="n", axes=F, xlab="", ylab="", xlim = c(0.5,6), ylim= c(0,8),
     main = "5 fold cross validation illustration")
for (i in 1:5){
  for (j in 1:5) {
    col <- ifelse(i==j, "red", "green")
    drawPoly(i,j, col)
  }
}

drawPoly(1,6, "red")
drawPoly(1,7, "green")

for(i in 1:5){
  text(i+0.5, 0.5, label = paste("k = ", i, sep = ""))
}

for(i in 1:5){
  text(0.5, i+0.25, label = 6-i)
}

text(3, 7.3, label = "Training Data")
text(3, 6.3, label = "Test Data")
```

